The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-17 trunk/tasbot/paper/paper.tex
About Super Mario

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  % XXX this can get cooler.
  {\bf Parameter Alert!}\,\, #1
}

@@

The basic idea is to deduce an objective function from a short recording of a player's inputs to the game. The objective function is then used to guide search over possible inputs, using an emulator. This allows the player's notion of progress to be generalized in order to produce novel gameplay. A design goal is that the objective function be amusingly elegant (not at all smart, fancy, or customized to the game) in order to demonstrate that the game is reducible to such a simple objective. The search needs to be game-agnostic and practical, but since the space is exponential ($256^{n}$), we need to be smart here.

@@

% (32kB of ROM is addressable at a time, and most games used bank switching to access hundreds of kilobytes of game included in the cartridge.)

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-global}
\end{center}\vspace{-0.1in}
\caption[lof not used]{
  A single maximal tight valid lexicographic ordering for my
  4,000-frame input training data to Super Mario Bros.. This function
  is globally nondecreasing, and is the decimal memory locations
  $\langle$\!~232, 57, 73, 74, 75, 115, 130, 155, 32, 184, 280,
  491, 506, 1280, 1281, 1282, 1283, 1288, 1290, 1337, 1338, 1339,
  1384, 1488, 1490, 1496, 1497, 1498, 1499, 1514, 1873, 1882, 1888,
  1904, 1872, 1906, 112, 113, 114, 2009, 2010, 2011, 1539~\!$\rangle$.

@@

\paragraph{Weighting.} To reduce the importance of randomness in the

@@

  defined when the memory is greater than all observed memories, which
  we will encounter later. The code returns $|V|/|V| = 1$ in that case,
  which is as good as anything.} This gives us the ability to compare
objectives on an absolute scale.\footnote{This is certainly not the
  only way to do it, and it has some questionable properties like
  ignoring the magnitude of change. But it is very simple.} Weighting
an objective is now simple:

@@

\begin{code}
for (;;) \{
  vector<uint8> before = GetMemory();
  vector<uint8> state = GetState();
\
  set<vector<uint8>> nexts;
  for (Future f : futures) \{
   nexts.insert(f.First10Frames());
   f.ChopOffFirst10Frames();
  \}
\ 
  while (nexts.size() < NFUTURES)
   nexts.push\_back(/* random motif */);
\ 
  for (vector<uint8> &next : nexts) \{
    RestoreState(state);
    for (uint8 i : next) Step(i);
    double score = ScoreByFutures(before, futures);
    // Save the best-scoring next...
    // 
  \}
\
  ReplaceBadFutures(futures);
  ExtendFuturesToDesiredLengths(futures);
\}
\end{code}

@@

The function {\tt ScoreByFutures} saves the state, then runs each
of the {\tt NFUTURES} futures to get a final state and memory. We
score each final memory relative to the start memory. Scoring is
a matter of

\subsection{Caching of emulation}

@@

\subsection{Backtracking}

\section{Results}

These games were all played with the same settings developed and tuned
for Super Mario Bros.; that is, this was the result of just running
the program the first time on a recording of me playing the game briefly.

@@

This is a very difficult game featuring a skateboarding cherubic
island guy called Master Higgins, whose girlfriend has been kidnapped
by King Quiller. He basically runs to the right and can throw or ride
stuff that he finds in eggs. The controls are pretty soft and danger
is everywhere. The objective functions that work are probably pretty
similar to Super Mario Bros., but there are fewer obstacles to
navigate---the difficulty for humans mostly comes from the speed and
reaction time. Master Higgins doesn't care about bumping into rocks
and dropping his health almost to nothing (but then is careful about
health), and once he gets a weapon his aim anticipates off-screen
enemies and he looks pretty savvy (Figure~\ref{fig:adventureisland}).
His weakness regards holes in the ground, which he sometimes jumps
into, probably for the same reason that Mario sometimes does. His
``pot bonus'' was 16720. Master Higgins beats several levels and makes
it into the cave before losing his last life by jumping straight into
a vibrating skull with swirling fireballs around it, which on a scale
of Darwin Award to dignified death is approximately a 6, in my
opinion.

@@

\subsection{Bubble Bobble}

>>>>

\newcommand\comment[1]{}
\newcommand\parameteralert[1]{
  \begin{quotation}
  \noindent {\bf Parameter Alert!} #1
  \end{quotation}
}

@@

The basic idea is to deduce an objective function from a short
recording of a player's inputs to the game. The objective function is
then used to guide search over possible inputs, using an emulator.
This allows the player's notion of progress to be generalized in order
to produce novel gameplay. A design goal is that the objective
function be amusingly elegant (not at all smart, fancy, or customized
to the game) in order to demonstrate that the game is reducible to
such a simple objective. The search needs to be game-agnostic and
practical, but since the space is exponential
($256^{n}$)\cite{GeneralizedSuper}, we need to be smart here.

@@

% (32kB of ROM is addressable at a time, and most games used bank
% switching to access hundreds of kilobytes of game included in the
% cartridge.)

@@

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario-global}
\end{center}\vspace{-0.1in}
\caption[lof not used]{
  A single maximal tight valid lexicographic ordering for my
  4,000-frame input training data to Super Mario Bros.. This function
  is globally nondecreasing, and is the decimal memory locations
  $\langle$\!~232, 57, 73, 74, 75, 115, 130, 155, 32, 184, 280,
  491, 506, 1280, 1281, 1282, 1283, 1288, 1290, 1337, 1338, 1339,
  1384, 1488, 1490, 1496, 1497, 1498, 1499, 1514, 1873, 1882, 1888,
  1904, 1872, 1906, 112, 113, 114, 2009, 2010, 2011, 1539~\!$\rangle$.

@@

\paragraph{Weighting.} \label{sec:objectiveweighting} 
To reduce the importance of randomness in the

@@

    defined when the memory is greater than all observed memories,
    which we will encounter later. The code returns $|V|/|V| = 1$ in
    that case, which is as good as anything.} This gives us the
  ability to compare objectives on an absolute scale.\footnote{This is
    certainly not the only way to do it, and it has some questionable
    properties like ignoring the magnitude of change. But it is very
    simple.} Weighting an objective is now simple:

@@

\begin{code}
for (;;) \{
  vector<uint8> before = GetMemory();
  vector<uint8> state = GetState();

  set<vector<uint8>> nexts;
  for (Future f : futures) \{
    nexts.insert(f.First10Frames());
    f.ChopOffFirst10Frames();
  \}

  while (nexts.size() < NFUTURES)
    nexts.push\_back(/* random motif */);

  for (vector<uint8> &next : nexts) \{
    RestoreState(state);
    for (uint8 i : next) Step(i);
    double score =
      ScoreByFutures(before, futures);
    // Save the best-scoring next...
  \}

  ReweightMotifs(best\_next, motifs);
  ReplaceBadFutures(futures);
  ExtendFuturesToDesiredLengths(futures);
\}
\end{code}

@@

The function {\tt ScoreByFutures} saves the state, then runs each
of the {\tt NFUTURES} futures to get a final state and memory. We
score each final memory relative to the start memory. The particulars
of scoring are not as interesting as the general idea, which is:
\begin{itemize}
\item The potential 10-frame {\tt next} sequence that we're
evaluating gets an {\em optimistic} score. This is based on the futures
for which the objectives {\em go up}. This is always non-negative.
\item Each future is also given a score, which is the sum of scores
from all the different {\tt next} sequences, regardless of their sign.
\end{itemize}

The purpose of the first is to identify a good {\tt next} sequence. We
take the sequence with the highest optimistic score. The idea is that
this is the sequence that gives us the best outcome if we continue
to control what we do in the future.

The purpose of the second is to identify which futures are good. A
good future tends to bring good outcomes regardless of the immediate
next step. Random futures that make is walk left when the good stuff
is to the right, or jump when there are spikes nearby above our head,
will receive negative scores for many or all of the {\tt next}
sequences.

{\tt ReweightMotifs} changes the weight of motifs that match the best
{\tt next} sequence that we just committed to, if we think that we
made progress on this step. The idea is to learn which sequences tend
to work well for us; this might be different from what the human
player did. For example, in run-to-the-right-and-jump kinds of games,
human players tend to hesitate more before obstacles and enemies than
a computer does.\footnote{That said, this part was an early idea and
  is probably not necessary. It's suspicious because these 10-frame
  sequences are not necessarily motifs (10 is the same as the normal
  motif length, and futures are generated by motifs, but they can
  become desynchronized because of the final incomplete motif, future
  mutation, and other things). So sometimes the chosen sequence
  doesn't affect weights. I think this would be better if we kept a
  Markov model } Knowing whether we made progress is kind of
difficult, since we can almost always find some move that increases
the objectives locally. For this I use the same approach of {\em value
  fraction} from Section~\ref{sec:objectiveweighting} based on a
sample of memories that we've seen so far. If we appear to be
progressing then the motif is weighted up; if we're regressing then it
is weighted down.

{\tt ReplaceBadFutures} kicks out the the futures with the worst total
scores, so that over time the random futures become good futures. Of
course, we always have to keep randomly adding to the end of each
future, and who knows what crap we'll find? A late addition to the
algorithm replaces some proportion of these with mutated versions of
the best scoring future. This helps us from messing up a good future
by appending crap to it, since we have multiple copies of it. It also
makes search more efficient, since futures that share common prefixes
can often benefit from caching (Section~\ref{sec:caching}). Currently
mutating a future involves chopping off its second half (it will get
extended to the desired length in the next step), and a \sfrac{1}{7}
chance of {\em dualizing} the entire sequence. Dualization swaps
the left and right buttons, the up and down buttons, B and A, and 
select and start. The idea of this is to occasionally introduce
very different futures to help us get out of stuck situations where
we should really be turning around or something like that.

During the development and tuning of the algorithm, I sometimes
observed {\tt ReweightMotifs} and {\tt ReplaceBadFutures} conspiring
to create a total monoculture of futures, where the weight of a single
motif (specifically ``press no buttons'') went out of control and all
futures just consisted of waiting. Waiting is usually a local
improvement to some objectives because of internal frame counters and
the cursor used to control music playback. To guard against this,
motifs have a maximum (and minimum) possible weight in {\tt
  ReweightMotifs}, and {\tt ReplaceBadFutures} always ensure that
some fraction of the futures are completely random (ignoring motif
weights).

\parameteralert{
  This is really the worst part in terms of parameters. They are: 
  {\tt NFUTURES}, the number of futures to maintain (40);
  {\tt NWEIGHTEDFUTURES}, the number of futures that are weighted,
  as opposed to totally random (35);
  {\tt DROPFUTURES}, the number of the worst-scoring futures we
  completely drop (5);
  {\tt MUTATEFUTURES}, the number of worst-scoring futures that
  we replace with mutants of the best future (7);
  {\tt MINFUTURELENGTH} and {\tt MAXFUTURELENGTH}, which put bounds
  on the sizes of futures (50 and 800);
  {\tt OBSERVE\_EVERY}, the frequency with which we sample memory
  for computing the value fraction for motif weighting (10);
  {\tt ALPHA}, the multiplicative factor for up- or down-weighting
  motifs (0.8);
  {\tt MOTIF\_MIN\_FRAC} and {\tt MOTIF\_MAX\_FRAC}, the bounds
  on motif weights (0.1 and 0.00001).

  Each is a scarlet letter of personal shame and future work is to
  eliminate them. In my opinion the least excusable are the bounds on
  future length, since these are related to what portion of time is
  ``interesting'' from a game perspective---for example, the max
  future length must exceed the number of frames that transpire after
  Mario dies but before he respawns, or the algorithm cannot properly
  avoid death. In my opinion this requires too much knowledge of the
  game. I don't have a good idea about how to compute this, at least
  without a human providing an input movie of dying (which would
  help for lots of other reasons)---but that would complicate the
  output of the learning process, which violates a primary design goal.

  {\tt NFUTURES} is another bothersome one, but there is more hope here.
  Basically it trades off computation time for quality of play, and
  more seems to always be better. We do have runtime metrics that could
  be used to dynamically change NFUTURES. For example, we could use
  the notion of improvability from Section~\ref{sec:backtracking}, or
  the value fraction, the gauge the marginal value of searching an
  additional future. Or something like that. This might actually help
  performance because we do the same amount of work (modulo caching)
  even during periods that the inputs are being ignored, like between
  worlds in Super Mario Bros., as we do when the going gets rough,
  and searching for the exact right delicate sequence of inputs would
  benefit from more options. The appearance of pausing in the output
  for Bubble Bobble (Section~\ref{sec:bubblebobble}) suggests that it
  knows all the futures are bad and needs to search different ones,
  and corroborates this idea.}

\subsection{Backtracking} \label{sec:backtracking}

The algorithm above always makes progress at the same rate of 10
frames per iteration. Sometimes it can get stuck in local maxima.
Super Mario Bros., my tuning game for this work, seems almost to have
been designed to thwart {\tt playfun}. The first level is pretty easy
once you have a decent algorithm, and early versions beat it without
any trouble. It's mainly about avoiding Goombas and planning far
enough ahead to make high jumps over pipes and pits (see the
accompanying videos for amusing struggles with these). World {\tt 1-2}
provides a challenge you may recognize from your youth, immediately
followed by something that computers find much harder
(Figure~\ref{fig:mario12}).

I literally spent about six weekends and thousands of hours of CPU on
this problem. First, the overhang is set up so that enemies emerge
from it just as you arrive. This means that loose play (imperfect
enemy avoidance) tends to get you killed right here. Mario has found a
number of different solutions to this problem, from waiting, to
kicking turtles from the left to clear out the enemies, to timing his
jump perfectly (it is possible!) to stomp the turtle after bouncing his
head off the ceiling. It's fun to see the different approaches and is
a good benchmark for whether the search algorithm is actually doing a
good job at tactics.

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario12}
\end{center}\vspace{-0.1in}
\caption{This is where it starts to get hard, even with lexicographic
ordering and time travel. This overhang is very tight in terms of the
inputs that can get you through it; random play tends to kill you because
of the advancing enemies that can't easily be jumped over. Greedy
algorithms can't resist the ledge with the coins, but then would need to
turn around.}
\label{fig:mario12}
\end{figure}

Immediately after this is an important test of longer-term planning.
Most of my early solutions would jump up on this ledge with 4 coins,
since the score is a component of many lexicographic
orderings\footnote{Maybe even all of them, since it should be globally
  obeyed; it's therefore a valid extension to any ordering that
  doesn't already include it. It's not necessarily a tight extension,
  however, so it may be excluded for that reason, or because during
  initialization it is not actually zero and so not globally obeyed. I
  never checked this stuff because I wanted to avoid any temptation to
  overfit the learning algorithm to the particulars of any game.} and
coins give you points. Down below is undesirable in the short term
because of the enemies. Then Mario would feel stuck and just hump up
against the wall jumping in this tiny dead end until time ran out. The
game contains some bugs (you have probably seen them) that allow the
screen to be scrolled without Mario actually progressing; these little
mini scrolls, plus the givens like the frame counter and music cursor,
prevent Mario from getting out of the local maximum. This is extremely
frustrating. I decided to add some longer-term planning into the
search procedure in order to try to help in these kinds of situations,
as well as the occasional deaths I would see.

Every {\tt CHECKPOINT\_EVERY} frames we save the state, and then every
{\tt TRY\_BACKTRACK\_EVERY} rounds, we do a backtracking phase if we
can. We just need to find a checkpoint at least {\tt
  MIN\_BACKTRACK\_DISTANCE} frames in the past. Call that point
{\tt start} and the current point {\tt now}. The backtracking
routine looks like this:

\begin{itemize}
  \item Let {\tt improveme} be the inputs between {\tt start} and {\tt now}.
  \item Get {\tt replacements}, a set of input sequences we might use
    instead. These are usually based on {\tt improveme}.
  \item Add {\tt improveme} to the set of replacements.
  \item Truncate the movie to {\tt start}.
  \item Now, do the normal playfun loop as though the (truncated) futures
    are whatever our current futures are, and the set of {\tt next}
    sequences are the {\tt replacements} array.
  \item Whatever the best one among those is, keep it. Don't update motifs
    or futures, however.
\end{itemize}

The idea is that we take a large sequence from our recent past, and the
same futures we're already using, and see if that sequence can be improved,
according to our objective functions, and using our same futures. Since
the existing sequence is always one of those, if it is the best one,
then the backtracking phase is a no-op. If we find something better, we
slot it in and continue. So the only risk here is if our objective functions
aren't good (we take as given that they are) and the only cost is time.

Generating the candidate set of replacements uses a bunch of different
techniques. They are:

\paragraph{Random.} We generate a random sequence of the same length as
the {\tt improveme} sequence.

\paragraph{Opposites.} We dualize (swap left with right, up with down, start 
with select, and B with A) and/or reverse random spans of the {\tt
  improveme} sequence. The idea is to try to introduce some variety in
cases where we're really getting stuck. This was specifically in hopes
that Mario would walk off the coin ledge in {\tt 1-2} and then find that the
course was clear below. I felt okay about this since this seems to be a
generic notion (the buttons do have semantics that are common to most NES
games where left and right are opposites), but that may just have been
rationalization since I was getting desperate. It didn't work; see below.

\paragraph{Ablation.} Randomly blanks out buttons from the input. For
example, if we don't need to jump and jumping is slowing us down or
something, this can remove that and make a small improvement to the
sequence.

\paragraph{Chop.} Removes random spans from the input. This iteratively
removes spans as long as the previous span was an improvement (see
below). This can cause the movie to get shorter by removing parts that
weren't contributing anything.\footnote{However, in games where an
  objective function includes something like a frame counter or the
  music cursor, shorter sequences always score lower than
  otherwise equivalent longer sequences.}

We use the following formula to score a potential improvement:

\begin{center}
\includegraphics[width=0.95 \linewidth]{isimprovement}
\end{center}

The {\tt start} state is as discussed, {\tt old\_end} is the state at
{\tt now}, and {\tt new\_end} is the state after the potential
improvement sequence. $\int_s^o$ is the integral of score changes
along the {\tt improveme} path and $\int_s^n$ along the candidate
improvement. The integral is the weighted sum of the objectives
increased minus the weighted sum of the objectives that decreased, at
each step, all summed up. This works better than just computing the
weighted score from e.g. {\tt start} to {\tt old\_end} when they are
separated by many frames (typical backtrack distances are several
hundred frames). The expression $n-o$\footnote{This is not numerical
  minus but minus according to the set of objective functions. Just
  roll with it.} is just the weighted sum of objectives that increased
minus the weighted sum of objectives that decreased between the old
end state and new end state; we have no option of computing the
integral here because we don't have an input sequence that goes from
the old end to the new end (and there almost certainly isn't one). We
require that three conditions hold:
\begin{enumerate}
\item $\int_s^n \geq \int_s^o$,
\item $\int_s^n > 0$
\item $n - o > 0$
\end{enumerate}

The integral has to be at least as good as before, the new integral
has to be positive, and the new end state needs to look better than
the old end state (the triangle inequality does not necessarily hold
here). If they all do, then the score for the improvement is $\int_s^n
- \int_s^o + (n - o)$, which is guaranteed to be positive.

Even if we have many possible replacements, we try only the ones with
the best scores; due to the particulars of the implementation there is
not a simple constant describing this maximum, but it's around 220
with the parameter settings used in this paper.

\parameteralert{
  Again! The humiliating appearance of constants! There are many here,
  having to do with the number of potential replacements we use of
  each type, the maximum number we keep, how often we bactrack, and
  how far. I am not totally satisfied with how backtracking works (see
  below), so I don't want to spend too much energy speculating about
  how to reduce the parameter space here; I'd rather replace the
  method wholesale.}

The {\em improvability} of a state is the fraction of these potential
improvements that are legal improvements, as above. Based on my
analysis, states that are highly improvable ($> 30\%$) tend to be
``worse'' (more stuck, closer to death, or whatever) than those that
are not very improvable ($< 5\%$). This isn't being used for anything
currently, but getting an absolute picture of how good a state is (as
opposed to simply comparing it to an adjacent state) is one of the
difficulties with this approach, so this may be a useful notion for
future directions.

\paragraph{Takes of woe, tales of joy.} 
Backtracking was initially added in order to fix the mario coin ledge
problem in {\tt 1-2}. It did not help with this. Older versions of the
code would have Mario get past this spot, probably by luck, but in
these he would get stuck in an embarrassing way on a small pit later,
or jump straight into an enemy. Most of the approaches that actually
looked solidly good elsewhere were failing badly here. As I started
writing up this paper, on an airplane, I discovered the bug in the
weighting of objective functions described in
Section~\ref{sec:objectiveweighting}. On my trip I had let {\tt
  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
lots of backtracking candidates, etc.) and it had spent about a
thousand CPU hours playing to this point, grabbing the coins, and then
dying of timeout, then losing all its lives, starting over, and
getting to this point again! After fixing the bug, I tried again, and
it was a huge breakthrough: Mario jumped up to get all the coins, and
then almost immediately jumped back down to continue the level! On his
first try he beat {\tt 1-2} and then immediately jumped in the pit at
the beginning of {\tt 1-3} just as I was starting to feel suuuuuper
smart (Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was
like {\em whaaaaaaat}?

\begin{figure}[hb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{mario13}
\end{center}\vspace{-0.1in}
\caption{After making me feel so successful by finally cleanly
  finishing world {\tt 1-2}, Mario immediately jumps into the first
  trivial pit in {\tt 1-3}. I believe what's going on here is that
  this is actually the best future, because resetting back a single
  screen isn't a large loss, lives aren't included in the objective
  function, and probably the rest of the futures were struggling to
  make progress. This level is very jumpy and probably needs more
  futures to navigate its obstacles. In any case, Mario, Touch\'e.}
\label{fig:mario13}
\end{figure}

Seeing this huge improvement changed my idea about what part of the
code needed work (I now believe that simpler search strategies might
work and better lexicographic order generation and weighting is called
for). But, this was already in the midst of writing up the paper, so
instead I spent the CPU time running the current version on a bunch of
games. Those results are described in Section~\ref{sec:results} and
videos are available at \url{http://tom7.org/mario/}. In the next
section I describe some of the performance improvements I made, and
other modifications to the emulator interface.

\section{Performance} \label{sec:performance}

Performance is really important, both because the quality of the
output is CPU-bound and because I am impatient. In its current state,
running {\tt playfun} is an overnight affair; it takes about an hour
of real time to produce 1000 output frames (16 seconds of gameplay) on
a high-end desktop machine. Other than a trick that I didn't end up
using, these performance ideas are not at all new, but they are
documented here for completeness. I certainly spent lots of time on 'em!

\subsection{Caching of emulation} \label{sec:caching}

The most expensive part, by far, is running a step of emulation. It
takes about 500\textmu s to emulate a single frame, though this can
vary depending on what instructions are actually executed (recall that
each frame corresponds to many many 2A03 instructions; there are
approximately 16,000 clock cycles per frame!). Hardware like the
Picture Processing Unit and sound chip are emulated as well, which may
actually be less efficient than the actual hardware (for example, the
sound hardware can implement filtering with passive physical
components like capacitors, but FCEUX contains a finite impulse
response filter running on IEEE floating point numbers). We want to
avoid executing a step more than once, if at all possible.

Caching is so canonical that some have called it (pejoratively) the
only idea in Systems research. And it is what we do here. The emulator
adds a call
%
\begin{code}
void CachingStep(uint8 input);
\end{code}
%
with exactly the same semantics as {\tt Step}. However, it first
checks to see if this exact input has been executed on this exact
start state before. If so, it just restores the cached result
state instead of executing the step.

I use the excellent CityHash function\cite{CityHash} as the hash code,
and use a least-recently-used approach to clean out the hash table
when it has more than a fixed {\em slop} amount more than the target
size. States are large because they contain both the savestate of the
before and after state. I have plenty of RAM, so I allow each process
to store up to 100,000 states with 10,000 states of slop, which takes
about 3 gigabytes per process.

Caching adds a little bit of overhead (a single step is about 13\%
slower, amortized), from saving the state, computing the hash code,
copying the state into the table, and cleaning the hash table
periodically. A cache hit is a zillion times faster than actually
executing the step, however, so as long as the hit rate is more than
13\%, it's worth doing. I use caching step in most places inside {\tt
  playfun}, except when I know that the state can't have been computed
yet. The {\tt playfun algorithm} is particularly suitable to caching:
The set of potential next input sequences {\tt nexts} usually share
several input prefixes, and we keep around futures for some time, so
we can often end up evaluating the same one under the same conditions
many times. Moreover, mutant futures usually share their first half,
making them half price. A really important property of caching is that
it's based on the state of memory but doesn't care about the actual
sequence used to get there. This means that in cases where the input
is ignored (Mario ignores the jump button most of the time while in
the air, for example, and ignores all inputs between levels) we reach
equivalent states and can use the cache if the next input is exactly
the same. The ideal approach here would be to follow how the bits of
the input are actually read by the code, and insert a more generic key
into the hash table. For example, if we see that the code never even
read the bit of the input corresponding to the up direction, then we
can reuse the step for an input that matches all bits but the up bit!
This of course requires a lot of mucking around in the internals,
which on a scale of within the scope to beyond the scope of this
article is a 9.9.

{\bf Software engineering lesson:} Initial results from caching were
disappointing and I thought the overhead was to blame. I made several
low-level tweaks to avoid copying, etc., to reduce the overhead from
36\% to 13\%. Then I discovered that there were 0 cache hits ever,
because of a bug (I was inadvertantly using pointer equality on keys
instead of value equality, so keys were never found). Check basic
things about your code's correctness before going on an optimization
sortie!

\subsection{Space optimizations}

Before I had the idea that is the actual subject of this paper, I was
playing around with other emulator search ideas that involved storing
very very large numbers of states. This necessitated minimal
representations of savestates. FCEUX uses zlib internally to compress
states, which works well; they shrink from something like 13,776
bytes\footnote{I don't completely understand the NES architecture and
  emulation strategy, but I believe some games include expansion chips
  that require more space for save states. All this work is based on
  the 2048 bytes of main NES RAM, as you already know. Perhaps clever
  video game authors from the past can protect against this paper's
  approach by storing important game facts inside expansion chips
  in the cartridge.} to an average of 2266. I made some modifications
to the save/load routines to make this as small as I could manage.
I removed the backing buffer, which is only used for drawing graphics,
movie info, header bytes that are constant or already known (savestate
size), which shrunk the compressed size to an average of 2047.42 bytes.
That meant I could fit about 32 million savestates in RAM at once, which
would be kind of amazing if I could tell my 8 year-old self that.

Compression algoritms work well when there is lots of redundancy in
their input. Comparing memories between two frames in some game,
you'll often see only a few differences. Some of the memory contains
stuff that's essentially read-only, even. To improve the
compressibility of savestates, I introduce the idea of a {\em basis}
state. This is any non-empty sequence of bytes which is an additional
argument to the {\tt LoadState} and {\tt SaveState} functions. It's
subtracted from the savestate prior to compression (it doesn't have to
be the same size; it's treated as an infinite repetition of those
bytes). If the basis is the same as the savestate, then the result is
all zeroes, which compresses nicely (of course, now you need to have
the basis in order to load the state, which is the same as the
savestate, so that didn't save you much!). But memories during
gameplay often don't differ much, so I save a state from any arbitrary
frame in the game and then use that as the basis for everything else.
This factors out any common parts and makes the states much more
compressible: The average size drops to 1870.41 bytes.

Using a better compression algorithm like the Burrows--Wheeler
transform\cite{Burrows}\footnote{This is really one of the all-time
  great algorithms. If you don't know it and you like this kind of
  thing, you should check it out. It's so surprising and elegant.}
would probably help too; zlib implements LZW which is only mediocre.
However, on a scale of throwing my computer out the window to wanting
to reimplement all software in my own private SML library, I just
can't stand getting 3\rd-party libraries to compile and link into my
applications, so I didn't even try. In any case, I abandoned this
direction, which was not working well even with lean savestates.

For {\tt playfun}, RAM is not a bottleneck at all. I disable
compression completely on savestates, including the builtin zlib stuff
(which is actually rather slow) and just use raw savestates. The
removal of unnecessary headers and stuff is still there, and saves a
few bytes and CPU cycles, but there was no reason to actually do it
for this particular work. {\em But sometimes I do unnecessary things.}

@@

The {\tt playfun} algorithm involves running 40 or so different
10-sequence {\tt next} steps and then scoring them against {\tt
  NFUTURES} different futures. Each {\tt next} and each future is
completely independent and dominated by the cost of evaluating
emulator steps. The ideal thing would be to use threads and shared
memory to run these steps in parallel. As I mentioned earlier,
FCEUX is hopelessly single-threaded.

Eventually I realized I needed to search more states than I could in a
single thread, and so I created MARIONET. It's a play on words, a
double entendre of ``Mario network'' and ``Marionette'', which is
obvious. This is a mode whereby {\tt playfun} can be started in
``helper'' mode (listens on a port for work to do) or ``master'' mode
(connects to a list of ports to run work on), in order to run
many emulators in different processes.

The processes communicate over TCP/IP. I only run it on my local
computer, but it would be reasonable to run it on a cluster or
possibly even the Internet. I used SDL's SDL\_Net\cite{SDLNet} as a
portable network interface in the hopes of keeping it
platform-agnostic (FCEUX is basically portable so it would be a shame
to make this Windows-specific, though I am not eager to try to get
this thing to compile on other platforms, I gotta be honest). SDL is a
nightmare to get working with mingw in a 64-bit compile, as usual, and
SDL\_Net contained some bugs\footnote{In particular, the {\tt
    SDLNet\_TCP\_Recv} call is supposed to block until it has received
  the entire requested length, but it occasionally returns early.}
that I had to spend some time working around. Anyway, I'm just
complaining. For serializing requests and responses to bytes, I used
Google's Protocol Buffer library\cite{ProtoBuf}, which is designed for
this kind of thing.

Helpers all start up with the same ROM and motif and objectives
loaded, so that they can simulate whatever the master would
compute.\footnote{Only the master reweights motifs, since it needs a
  single sample of memories that we've observed. That means that in
  cases where a helper generates random motifs, it does so with
  respect to the original human weighting. There are some other small
  things I simply avoided doing because they would require too much
  coordination between processes or make the RPCs too large; MARIONET
  was part of the system for much of its development.} They just wait
on a port for the master to send a request, which can be either ``try
this set of {\tt nexts} with these futures and send me the results'' or
``try to find $n$ improvements of these inputs {\tt improveme} using
so-and-so approach.'' In either case we send the base state as a
serialized savestate.

The master does the outer loop of the algorithm and all the writing to
disk and so on. When it needs to do an expensive step like the inner
loop, it prepares a set of jobs and sends them to workers using a
little fork-join library ({\tt netutil.*}). The library keeps track of
the outstanding jobs and which helpers are working, and feeds them
work when they are idle, until all the jobs are complete. It's even
got a color ANSI progress bar and stuff like that.

Utilization is very good (Figure~\ref{fig:utilization}), and we get
almost a linear speedup with the number of processes, up to the number
of hardware threads the system supports (twelve). Actually, in
addition to the computer remaining usable for other stuff, $n-1$
helpers may work a bit better than $n$, because each of the
helpers is then able to avoid preemption. Since the driver loop is
synchronous, having one laggard process slows the whole darn thing
down while everything waits for it to finish.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.75 \linewidth]{utilization}
\end{center}\vspace{-0.1in}
\caption{Utilization with 12 helpers and one master on a 6-core (12
  hyperthreads) processor. Looks good. Keeps the bedroom warm.}
\label{fig:utilization}
\end{figure}

MARIONET is a huge improvement and was what enabled using 40+ futures
in a reasonable amount of time, which is key to high-quality output. It's
obviously the way you want to run the software, if you can get it to
compile. There is one downside, though, which is that it reduces our
ability to get lucky cache hits across {\tt nexts} that are similar (or
have the same effect), and when replaying the winner as we commit to it.
It's worth it, but smartly deciding which helper gets which tasks (because
they share prefixes, for example) would save time. Running it all in one
process with a shared memory cache would be the best, as long as the
lock contention could be kept under control.

\section{Results} \label{sec:results}

In this section I describe the results of running {\tt learnfun} and
{\tt playfun} on several NES games. These games were all played with
the same settings developed and tuned for Super Mario Bros.; that is,
this was the result of just running the program the first time on a
recording of me playing the game briefly. Every game I tried is
documented here; lack of CPU time before conference is the main reason
your favorite is not included. The website \url{http://tom7.org/mario/}
contains video versions of some of these, which may be more fun. Let's
begin with the {\em ur} game, Super Mario Bros..

\subsection{Super Mario Bros.}

This game is an obvious benchmark and the one I used during the
development of {\tt learnfun} and {\tt playfun}. Some parts of the
algorithm (see e.g. Section~\ref{sec:backtracking}) were developed
specifically to get around problems in the ability to play this game,
though I believe simpler methods would also work on this game, now
that some important bugs are fixed.

Automated Mario is fun to watch, and definitely my favorite. The
familiarity of the game, the combination of human-like maneuvers and
completely bizarre stuff, daredevil play, and bug exploitation
(Figure~\ref{fig:mario13bug1}) are all a delight. It's even funny when
it fails, watching Mario struggling with obstacles like tiny pits, or
making a heroic move followed by a trivial mistake. I recommend
watching the videos.

This algorithm works well on Super Mario Bros., and I think that with
some improvements it could play quite far into the game. It probably
can't beat the game; Worlds {\tt 7-4} and {\tt 8-4} require some weird
puzzling that we all needed {\em Nintendo Power Magazine}'s help to
solve as kids, and are not just a matter of running to the right and
avoiding obstacles. In the current incarnation, Mario breezes through
World {\tt 1-1} every time, consistently beats {\tt 1-2} (I've only
tried about three times with the newest version of the code, but each
time he's succeeded) and has limited success on {\tt 1-3} but
eventually suicides too many times. I'll keep trying.

\begin{figure}[h!tb]
\begin{center}
\includegraphics[width=0.75 \linewidth]{mario13bug1a} \\[0.3em]
\includegraphics[width=0.75 \linewidth]{mario13bug1b}
\end{center}\vspace{-0.1in}
\caption{Mario bounces off one Goomba up into the feet of another. Not
  only doesn't he have enough velocity to reach the platform, but he's
  about to hit that Goomba from below. Believe it or not, this ends
  well: {\tt playfun} is happy to exploit bugs in the game; in this case,
  that whenever Mario is moving downward (his jump crests just before the
  Goomba hits him) this counts as ``stomping'' on an enemy, even if that
  enemy is above him. The additional bounce from this Goomba also allows
  him to make it up to the platform, which he wouldn't have otherwise!
}
\label{fig:mario13bug1}
\end{figure}

The Mortal Kombat-style {\em Finish Him!} to any machine learning
algorithm is overfitting, however. Does the technique work on other
games, without endless tweaking as I did with Super Mario Bros.? On a
scale of no to yes, this is both a 1 and a 10; some games work even
better than Mario and some are a disaster.

@@

This is a bad, difficult, but likable game featuring a skateboarding
cherubic island guy called Master Higgins, whose girlfriend has been
kidnapped by King Quiller. He basically runs to the right and can
throw or ride stuff that he finds in eggs. The controls are pretty
soft and danger is everywhere. The objective functions that work are
probably pretty similar to Super Mario Bros., but there are fewer
obstacles to navigate---the difficulty for humans mostly comes from
the speed and reaction time. Master Higgins doesn't care about bumping
into rocks and dropping his health almost to nothing (but then is
careful about health), and once he gets a weapon his aim anticipates
off-screen enemies and he looks pretty savvy
(Figure~\ref{fig:adventureisland}). His weakness regards holes in the
ground, which he sometimes jumps into, probably for the same reason
that Mario sometimes does. His ``pot bonus'' was 16720. Master Higgins
beats several levels and makes it into the cave before losing his last
life by jumping straight into a vibrating skull with swirling
fireballs around it, which on a scale of Darwin Award to dignified
demise is approximately a 6, in my opinion.

@@

\subsection{Karate Kid, The} \label{sec:karate}
